First steps towards a formalization of Forcing

### 2018-08-15 LSFA/rs.tex
DC vs Pointed_DC

\subsection{A sequence version of $\DC$}\label{sec:sequence-version-dc}
In order to prove the Rasiowa-Sikorski lemma in
general, a version of the Axiom of Choice ($\AC$) must be proved. This
version is known as the \emph{Principle of Dependent Choices}:
\begin{quote}
  ($\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
  $\forall x\in A.\,  \exists y\in A.\, x\rr y$, then there exists
  $f:\om\to A$ such that $f(0)=a$ and $f(n)\rr f(n+1)$ for all
  $n\in\om$.
\end{quote}

A different version of $\DC$ (without the constraint $f(0)=a$, but
allowing a parameter that permits to prove an equivalence with $\AC$)
was formalized by Krzysztof
Grabczewski~\cite{DBLP:journals/jar/PaulsonG96} as part of the
Isabelle/ZF library, along with many statements that follow from or
are equivalent to $\AC$. Since the proofs there required a greater
generality, it would have been very contrived to derive this
``pointed'' version of $\DC$ from that development. Instead, we
preferred to give a direct proof based on the one in
Moschovakis' notes \cite{moschovakis1994notes}. Another advantage is that
this proof depends on the following version of $\AC$ (existence of
choice functions on $\P(A)\sm \{\mty\}$):
\[
\exists (s: \P(A)\sm \{\mty\}\to A). \, \forall X\sbq A.\ X\neq \mty \limp
s(X) \in X.
\]
This statement is more standard than $\AC$ as stated in the
Isabelle/ZF development, which involves a silent use of Replacement
since it is written as an axiom scheme.

The strategy is to define the function $f$ discussed above by
primitive recursion on the naturals, which can be done easily thanks
to the package of Isabelle/ZF
\cite{paulson1995set,paulson2000fixedpoint} for definitions by
recursion on inductively defined sets.\footnote{The package figures
  out the inductive set at hand and checks that the recursive
  definition makes sense; for example, it rejects definitions with a
  missing case.}

@@

The proof is mostly routine; after a few lemmas we obtain the
following theorem:

>>>>

\subsection{A sequence version of Dependent
  Choices}\label{sec:sequence-version-dc} 
%% In order to prove the Rasiowa-Sikorski lemma in
%% general, a consequence of the Axiom of Choice ($\AC$) must be invoked. This
%% consequence is known as the \emph{Principle of Dependent Choices}:
The Rasiowa-Sikorski lemma follows naturally from a ``pointed''
version of   the \emph{Principle of Dependent Choices} which,
in turn, is a consequence of the Axiom of Choice ($\AC$). It is
therefore natural to take as a starting point the theory \isatt{AC}
which adds the latter axiom to the toolkit of Isabelle/ZF.

The statement we are interested in is the following:
\begin{quote}
  (Pointed $\DC$) Let $\rr$ be a binary relation on $A$, and $a\in A$. If
  $\forall x\in A.\,  \exists y\in A.\, x\rr y$, then there exists
  $f:\om\to A$ such that $f(0)=a$ and $f(n)\rr f(n+1)$ for all
  $n\in\om$.
\end{quote}

Two different versions of $\DC$ (called $\DC0$ and $\DC(\kappa)$) have
already been formalized  by Krzysztof 
Grabczewski~\cite{DBLP:journals/jar/PaulsonG96}, as part of a
study of equivalents to $\AC$ (following Rubin and Rubin
\cite{rubin1985equivalents}). Nevertheless, those are not convenient
for our purposes for the reasons we now explain. The axiom $\DC0$
corresponds essentially to our Pointed $\DC$ but without the
constraint $f(0)=a$; it is a nice exercise to show that $\DC0$ implies
Pointed $\DC$, but a formalization would have a moderate length. On
the other hand, the shortest path from $\AC$ as stated in the
corresponding Isabelle theory to $\DC0$ using the already formalized
material involves a complicated detour (130+ proof commands spanning
various files of the \isatt{ZF-AC} theory), compared to the mere 11 commands from $\AC$ to
\isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow}. This last
one is the
choice principle that we use in our formalization of Pointed $\DC$, and states 
the existence of
choice functions on $\P(A)\sm \{\mty\}$):
\[
\exists (s: \P(A)\sm \{\mty\}\to A). \, \forall X\sbq A.\ X\neq \mty \limp
s(X) \in X.
\]
Another advantage of taking
\isatt{AC{\isacharunderscore}func{\isacharunderscore}Pow} as a
starting point is that it does not involve proper classes: The version
of  $\AC$ in Isabelle/ZF corresponds to an axiom scheme of first order
logic and as such is not a standard formulation. 
%% 26 commands to get Pointed $\DC$.

The strategy to prove Pointed $\DC$ (following a proof in Moschovakis
\cite{moschovakis1994notes}) is to define the function $f$ discussed above by
primitive recursion on the naturals, which can be done easily thanks
to the package of Isabelle/ZF
\cite{paulson1995set,paulson2000fixedpoint} for definitions by
recursion on inductively defined sets.\footnote{The package figures
  out the inductive set at hand and checks that the recursive
  definition makes sense; for example, it rejects definitions with a
  missing case.}

@@

The proof is mostly routine; after a few lemmas (having 26 proof
commands in total) we obtain the
following theorem:
